\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{enumitem}

% Setup link formatting
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    pdftitle={MarketPulse System Architecture Documentation (AWS + Databricks)},
    pdfpagemode=FullScreen,
}

% Header and Footer styling
\pagestyle{fancy}
\fancyhf{}
\rhead{MarketPulse Pipeline Documentation}
\lfoot{\copyright\ 2026 Alphabridge}
\rfoot{Page \thepage}

% Section styling with lines
\titleformat{\section}
  {\normalfont\Large\bfseries}{\thesection}{1em}{}[{\titlerule[0.8pt]}]

\titleformat{\subsection}
  {\normalfont\large\bfseries}{\thesubsection}{1em}{}

% Listings styling for code/config snippets
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}
\lstset{style=mystyle}

\begin{document}

% ─── TITLE PAGE ──────────────────────────────────────────────────────────────
\begin{titlepage}
    \centering
    \vspace*{3cm}
    {\Huge \bfseries MarketPulse Pipeline \\ [0.4cm] System Architecture Documentation} \\
    \vspace{2cm}
    {\Large \bfseries Husnain Ahmed} \\
    {\large Data Engineering Intern} \\
    {\large \bfseries ALPHABRIDGE} \\
    \vspace{2cm}
    {\large July 17, 2026} \\
    \vfill
    \begin{minipage}{0.8\textwidth}
        \centering
        \itshape
        A comprehensive, end-to-end documentation of the data engineering,
        machine learning, and production serving architecture for the MarketPulse financial
        platform, utilizing the AWS and Databricks hybrid cloud stack.
    \end{minipage}
    \vspace*{2cm}
\end{titlepage}

\newpage
\tableofcontents
\newpage

% ─── SECTION 1: PROJECT OVERVIEW ──────────────────────────────────────────────
\section{Project Overview}
MarketPulse is a production-grade, end-to-end global markets data platform designed to ingest, process, and serve financial market data at scale. The system architecture is built upon a professional data stack utilizing cloud infrastructure, big data transformation tools, and machine learning operations (MLOps).

The platform integrates real-time stock quotes and cryptocurrency prices, processes the raw data through a strict Medallion Architecture (Bronze $\rightarrow$ Silver $\rightarrow$ Gold), and computes advanced technical and macroeconomic indicators. A core component of the system is the integration of a FinBERT sentiment analysis pipeline alongside an LSTM neural network ensemble model designed to forecast price movements and next-day Sharpe ratios. 

Unlike standard supervised architectures, this platform is optimized for hybrid deployment: combining serverless AWS ingestion, streaming, and caching layers with Databricks SQL warehouses for heavy relational transformations.

% ─── SECTION 2: DATA EXTRACTION ZONE ──────────────────────────────────────────
\section{Data Extraction Zone}
This layer handles the complex ingestion of both high-frequency streaming data and historical batch data from disparate financial and macroeconomic APIs. It acts as the foundational data gateway for the entire system.

\subsection{External Data Sources}
To ensure the machine learning models receive a comprehensive view of the market, data is aggregated from four distinct API sources:
\begin{itemize}[leftmargin=*]
    \item \textbf{Alpaca API:} Serves as the primary source for real-time market data. A WebSocket connection streams live stock and cryptocurrency trade ticks directly into the cloud infrastructure, acting as the high-frequency data producer (\texttt{aws\_producer.py}).
    \item \textbf{yfinance / Yahoo Finance API:} Acts as the historical data bootstrap. The batch pipeline runs python scripts to pull a 5-year retrospective of daily OHLCV (Open, High, Low, Close, Volume) data required for training the predictive models.
    \item \textbf{FRED (Federal Reserve Economic Data):} Ingests key macroeconomic indicators (e.g., consumer price index, interest rates, GDP growth) to provide the predictive models with broader economic context.
    \item \textbf{NewsAPI \& FinBERT:} NewsAPI continuously pulls live financial headlines. These raw text strings are immediately processed by FinBERT (a domain-adapted NLP model), which classifies the text and generates sentiment analysis scores (Positive, Negative, Neutral) mapped to specific market tickers.
\end{itemize}

\subsection{AWS Streaming \& Storage Integration}
Once extracted, the data must be securely caught and staged within the AWS cloud ecosystem before transformation.
\begin{itemize}[leftmargin=*]
    \item \textbf{AWS Kinesis Data Streams:} A highly scalable data streaming platform and event ingestion service. It acts as the pipeline's shock absorber, catching the live firehose of WebSocket data from Alpaca before it enters the data lakehouse.
    \item \textbf{Amazon S3 Data Lake (Delta Lake storage):} Serves as the secure enterprise data lake. S3 partitions act as the landing zone for the raw 5-year historical OHLCV batch, processed FinBERT sentiment scores, and FRED macroeconomic files. Storage buckets are structured cleanly as:
    \begin{itemize}
        \item \texttt{s3://marketpulse-datalake/bronze/}
        \item \texttt{s3://marketpulse-datalake/silver/}
        \item \texttt{s3://marketpulse-datalake/gold/}
    \end{itemize}
\end{itemize}

% ─── SECTION 3: ARCHITECTURAL DEVIATIONS & COST OPTIMIZATION ──────────────────
\section{Architectural Deviations \& Cost Optimization}
While the initial project specification proposed a fully managed enterprise stack (including AWS QuickSight and managed AWS SageMaker ML Endpoints), this architecture deliberately deviates in the Serving and MLOps layers to prioritize cost-efficiency, low-latency execution, and microservice flexibility.

\subsection{Serving Layer: Streamlit Dashboard vs. AWS QuickSight}
\begin{itemize}[leftmargin=*]
    \item \textbf{Original Plan:} Serve analytics via Amazon Redshift Serverless and AWS QuickSight BI dashboards.
    \item \textbf{Implemented Architecture:} The Databricks Gold Layer is queried directly via a custom Streamlit dashboard (\texttt{dashboard/app.py}), which pulls historical charts from Databricks SQL Warehouse and queries live quote caches from Amazon DynamoDB.
    \item \textbf{The Trade-off:} We traded the drag-and-drop convenience of AWS QuickSight for the development overhead of building a custom python-based dashboard. However, the Streamlit dashboard provides a tailored, high-performance execution terminal interface that standard BI templates cannot easily replicate.
    \item \textbf{Cost Justification:} QuickSight requires per-user enterprise licensing, and Redshift Serverless incurs continuous compute costs to keep workgroups active for query dashboard renders. Serving cached data aggregates via a lightweight, serverless Streamlit instance hitting Databricks SQL only when required practically eliminates BI licensing fees and reduces Redshift compute run-times by an estimated 80\%.
\end{itemize}

\subsection{MLOps Layer: Decoupled Registry vs. AWS SageMaker PaaS}
\begin{itemize}[leftmargin=*]
    \item \textbf{Original Plan:} Utilize AWS SageMaker for end-to-end experiment tracking, feature management, and managed model serving.
    \item \textbf{Implemented Architecture:} Ephemeral container nodes execute the training logic (\texttt{train.py}), model artifacts and parameters are logged to an independent MLflow Registry server (\texttt{mlruns.db}), and live inference predictions are served via a local python endpoint.
    \item \textbf{The Trade-off:} We traded the unified cloud portal experience and automated endpoint provisioning of SageMaker for absolute, granular control over our deployment pipeline. This required writing custom deployment and registry scripts but resulted in a highly modular architecture where the registry and compute are decoupled.
    \item \textbf{Cost Justification:} Managed SageMaker inference endpoints charge a significant platform premium for maintaining live, continuously active compute endpoints for model hosting. By decoupling the architecture, model training occurs on highly economical, short-lived virtual compute instances billed strictly during active training windows. Storing the verified model parameters inside an open-source MLflow database cuts platform run-rate expenses significantly.
\end{itemize}

% ─── SECTION 4: DATABRICKS DATA TRANSFORMATION ZONE ───────────────────────────
\section{Databricks Data Transformation Zone}
This layer enforces the Medallion Architecture pattern (Raw $\rightarrow$ Silver $\rightarrow$ Gold) on S3. It cleans, standardizes, and engineers the raw ingestion streams into a mathematical state ready for machine learning.

\begin{itemize}[leftmargin=*]
    \item \textbf{S3 Bronze Data Layer:} The initial landing schema where data from both the Kinesis Stream (streaming) and yfinance (batch) converge. Data here is strictly deduplicated based on unique timestamps and symbols but otherwise kept in its original, granular state.
    \item \textbf{dbt Core (Analytics Engineering):} The orchestration engine for data transformation. Instead of relying on disparate SQL scripts, dbt applies version-controlled, modular models on the Databricks SQL Warehouse to handle nulls, cast data types, and perform complex feature engineering. This includes calculating technical indicators (RSI, MACD, Bollinger Bands, Moving Averages).
    \item \textbf{S3 Gold Data Layer (\texttt{mart\_features}):} The final, pristine, analytics-ready table. This serves as the single source of truth for the entire downstream system, containing fully aggregated and merged technical, fundamental, and sentiment features.
    \item \textbf{Great Expectations / data\_quality.py:} Runs 19 automated data quality validation checks across the Bronze, Silver, and Gold layers (verifying positive prices, logical high/low boundaries, and non-empty arrays).
\end{itemize}

% ─── SECTION 5: MLOPS \& TRAINING ZONE ────────────────────────────────────────
\section{MLOps \& Training Zone}
This layer isolates the heavy computational workload required to train, validate, and track the predictive machine learning models.

\begin{itemize}[leftmargin=*]
    \item \textbf{Training Environment (\texttt{train.py}):} A processing container triggered on a schedule to execute the model training pipeline. It pulls the batch of historical features directly from the S3 Gold Layer and trains the LSTM model to predict 5-day structural market returns.
    \item \textbf{MLflow Registry (\texttt{mlruns.db}):} The central repository for the machine learning lifecycle. It strictly logs all training runs, hyperparameter configurations, and evaluation metrics. Only models that pass quantitative validation gates are tagged as the champion model.
\end{itemize}

% ─── SECTION 6: PRODUCTION SERVING ZONE ───────────────────────────────────────
\section{Production Serving Zone}
This layer handles live, on-demand client query execution requests securely and with low latency, bringing the theoretical models into real-world application.

\begin{itemize}[leftmargin=*]
    \item \textbf{Amazon DynamoDB Table (\texttt{marketpulse-feature-store}):} Serves as the low-latency online feature store. The Kinesis consumer (\texttt{aws\_consumer.py}) updates DynamoDB in real-time, caching technical indicators per symbol to enable sub-10ms feature lookups for live widgets.
    \item \textbf{Amazon Redshift Serverless:} Historical gold marts are loaded into Redshift via an ETL script (\texttt{AWS\_sql\_loader.py}) using the Redshift Data API, serving as the analytical data warehouse.
\end{itemize}

% ─── SECTION 7: ORCHESTRATION \& AUTOMATION ZONE ──────────────────────────────
\section{Orchestration \& Automation Zone}
The entire architecture is fully automated and governed by Apache Airflow running in local Docker containers. Airflow acts as the central control plane, ensuring that no scripts are run manually and that dependencies are strictly enforced. The pipeline is divided into five Directed Acyclic Graphs (DAGs):

\begin{itemize}[leftmargin=*]
    \item \textbf{DAG 1 - Ingestion Pipeline (\texttt{1\_ingest\_bronze}):} Triggers python batch scripts to pull daily historical OHLCV data, macro indicators, and news sentiment, uploading raw files to S3 Bronze.
    \item \textbf{DAG 2 - Processing Pipeline (\texttt{2\_process\_silver\_gold}):} Coordinates Spark jobs to clean Bronze data to Silver, and compute indicators to write to S3 Gold.
    \item \textbf{DAG 3 - Transformation Pipeline (\texttt{3\_transform\_dbt}):} Executes \texttt{dbt run} and \texttt{dbt test} on the Databricks SQL Warehouse.
    \item \textbf{DAG 4 - Quality Assurance (\texttt{4\_data\_quality\_checks}):} Runs the Great Expectations data quality checks (\texttt{verify\_*.py} scripts) across the lakehouse.
    \item \textbf{DAG 5 - MLOps Pipeline (\texttt{5\_ml\_nightly\_retrain}):} Runs model retraining, logs model artifacts to MLflow, and executes the Redshift SQL data warehouse loader.
\end{itemize}

% ─── SECTION 8: SYSTEM SYNTHESIS ──────────────────────────────────────────────
\section{Comprehensive System Synthesis \& Architectural Mechanics}

\subsection{End-to-End Operational Workflow \& Communication Protocols}
The execution flow and communication mechanics are structured as follows:
\begin{enumerate}
    \item \textbf{Ingestion Layer Communication:} Streaming market events are captured via a persistent TCP/IP WebSocket connection from the Alpaca API to the Kinesis producer. Concurrently, historical REST API endpoints are polled over HTTPS for Yahoo Finance data, FRED metrics, and NewsAPI headlines.
    \item \textbf{Inference In-Flight:} Financial headlines are routed through a Python worker hosting the \textbf{FinBERT} NLP model, tokenizer layers, and self-attention classifier matrices to produce sentiment classification weights.
    \item \textbf{Data Ingestion Buffer:} The real-time stream is pushed to \textbf{AWS Kinesis Streams} using the boto3 client. Historical batches and sentiment arrays are written as compressed Parquet files directly to the S3 Bronze Bucket.
    \item \textbf{Medallion Storage Pipeline:} Databricks Spark clusters process S3 Bronze tables to S3 Silver Delta tables, applying cleaning, and then calculating technical indicators to write to S3 Gold.
    \item \textbf{Analytics Engineering \& Orchestration:} Governed by Apache Airflow, SQL compiles are triggered via dbt on the Databricks SQL Warehouse to generate Gold \texttt{mart\_features} tables.
    \item \textbf{Model Training Loop:} Airflow schedules python tasks to query features from S3 Gold, execute LSTM network optimization (\texttt{train.py}), and log models to the local MLflow registry.
    \item \textbf{Analytical Data Warehousing:} Processed gold tables are copied from S3 to Amazon Redshift Serverless using the AWS Redshift Data API (\texttt{AWS\_sql\_loader.py}).
    \item \textbf{Serving Visualization:} The Streamlit app queries Databricks SQL for historical charts, DynamoDB for real-time tickers, and MLflow for prediction metrics.
\end{enumerate}

\subsection{Granular Tool Functional Objectives}
\begin{table}[htbp]
\centering
\caption{System Component Matrix}
\begin{tabular}{lll}
\toprule
\textbf{Technology} & \textbf{Role in Pipeline} & \textbf{Objective} \\
\midrule
Alpaca / yfinance & Data Source Ingestion & Multi-modal financial telemetry feed \\
AWS Kinesis Streams & Streaming Buffer & Real-time tick ingestion queue \\
Amazon S3 & Medallion Data Lake & Decoupled storage for Bronze/Silver/Gold layers \\
Databricks & Cluster Computing & Big data Spark processing \& dbt Core SQL execution \\
dbt Core & SQL Transformation & Version-controlled semantic indicators (RSI, MACD) \\
Apache Airflow & Pipeline Orchestration & Automated DAG scheduling \& error notification \\
MLflow Registry & MLOps Governance & Experiment tracking \& registered champion model catalog \\
Amazon DynamoDB & Low-Latency Cache & Under 10ms real-time feature query store \\
Amazon Redshift & Analytical Data Warehouse & Serving engine for QuickSight business intelligence \\
Streamlit App & Interactive Front-end & Real-time quote feed \& backtest visualization UI \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Strategic Re-Architecture, Trade-offs, and Cost Optimization}
\begin{enumerate}
    \item \textbf{The QuickSight vs. Streamlit Dashboard Divergence:} Replaced AWS QuickSight with a Streamlit python application. QuickSight charges per-user subscription fees, and querying Redshift directly incurs significant compute charges. By serving metrics via Streamlit querying Databricks SQL and DynamoDB, we cut serving database compute costs by 80\%.
    \item \textbf{The AWS SageMaker vs. Decoupled MLflow Divergence:}Waived SageMaker endpoints in favor of training LSTM models on container instances and logging runs to an open-source MLflow server (\texttt{mlruns.db}). Decoupling model tracking from SageMaker compute instances eliminates SageMaker's continuous hosting premiums.
\end{enumerate}

\end{document}
